Transactional Memory Concurrency Verification with Landslide

### 2018-03-15 htm/paper.tex
finished, minus proofreading

			\hline
			{\tt map\_basic}
			& 2,1 & \ETAdag{10d 17h} & \ETAdag{16388977} \\
			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
			\hline
			{\tt map\_basicer}
			& 2,1 & \cpu{877.44}& \ints{28635}	\\
			& 2,2 & \ETAdag{2d 7h} & \ETAdag{5925634} \\
			& 3,1 & 	& 	\\ % TODO
			% ^^ running now, certified not recompiled
			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified (or not) by Landslide.}
	\label{tab:verifs}
\end{table}

Several larger tests we were unable to complete before the submission deadline are also shown.
%Their completion times are listed as {\em unk.}
These are indicated with $\dagger$ %\ETAdag{}
and their listed ETAs and state space sizes represent Landslide's estimate %at the moment of timeout.
after a timeout of 1 hour
(hopefully enough for data-race PPs to saturate and estimates to stablize,
although note the two estimate types use different algorithms so may disagree significantly).
%For these, Table~\ref{tab:timeouts} presents their size estimates at the moment of timeout.
Though these tests proved beyond our reach,
in contrast with the high speed expected of bug-finding results,
%we believe
the verification guarantee may in some cases be worth the estimated time requirement
for widely-used industrial implementations.
Future work may also expand our coverage upon this frontier \cite{landslide-phdthesis}.

@@

We noticed significant slowdowns when varying several factors of our experimental environment.
For example, multiple Landslides running at once on our test machine slow each other down,
%(including multiple single state spaces in a single test when not using {\tt -M}),
likely arising from kernel resource contention as Landslide uses {\tt fork()} to save simulation state.
Table~\ref{tab:variants} shows the impact of running a single Landslide instance with {\tt -M} on
{\tt counter}(2,2) (384 distinct interleavings)
with various programs also running, despite never saturating our test system's 8 CPUs.
% for i in `seq 1 10`; do ./landslide -X -v -p counter -M 2>&1 | grep 'total CPU time'; done

Table~\ref{tab:variants2} shows
the impact of Landslide needing to automatically annotate {\tt counter}(2,2) being run for the first time,
rather than reusing existing instrumentation,
as well as the variance of Quicksand mode. %(i.e., without {\tt -M}).
%(2128 interleavings total).
The {\sf quicksand} variance was surprisingly bimodal,
with 6 samples in a distribution of 106.14$\pm$1.22 and 4 in 142.11$\pm$1.67,
suggesting two distinct scheduling patterns for its workqueue threads.
Future work should figure out why.

@@

Many such branches may be equivalent;
for example,
explicit and conflict aborts need not be tested separately in transactions
whose failure paths do not distinguish the cause,
and the perennial {\tt \_XABORT\_RETRY} abort can be skipped entirely
if the client abstracts it away behind a retry loop.%
\footnote{Proof left to future work. We've proved enough here already.}
In fact, applying both reductions simultaneously
produces a state space smaller than the original;
testing them by hand
\footnote{Using visual inspection of the test to trust the reductions' soundness.}
reduced {\tt map\_basicer}(2,1)'s 28635 interleavings to 11577.
% TODO: try map_basic and get a 1 hour estimate
%(although {\tt map\_basic}(2,1) lies still beyond reach).
%We tested both reductions by hand, trusting their soundness by visual inspection,
%on {\tt map\_basic}(2,1),
%and preview the resulting state space sizes in Table~\ref{tab:abort-codes}.
%
%\begin{table}[h]
%	\caption{...}
%	\label{tab:abort-codes}
%\end{table}
%
Future work could use static or dynamic flow analysis to
identify such reduction opportunities automatically;
for now, this feature is disabled by default but accessible via the {\tt -A} command-line option
to Landslide (in addition to {\tt -X}).

{\bf Relaxed memory orderings.}
Section~\ref{sec:design}'s formalization of thread interleavings does not account for read/write reorderings
possible on relaxed consistency architectures \cite{memory-consistency-models}.
In fact,
even after \cite{htm-mario}'s proposed fix,
our running example program is still incorrect on Total Store Ordering (TSO) architectures such as x86.
Despite stores being totally-ordered, x86 may still reorder stores after subsequent loads.
Accordingly, an execution of $8,9a,9b,9c$ in Figure~\ref{fig:htm-fixed}
may be locally visible to another thread as $9a,8,9b,9c$,
and hence an apparent interleaving of
\[
	\tiat{1},\tjat{1-5},\tiat{7},\underline{\tiat{9a}},\tkat{1-5},\underline{\tiat{8}},\tiat{9b-B}
\]
is possible
(reordered accesses underlined for emphasis).
An {\tt mfence} barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}.
On Partial Store Ordering (PSO) architectures, even more barriers may be necessary.

@@

We thank Mario Dehesa-Azuara for generously providing the HTM data structures and benchmarks used in our evaluation.
%We also thank the member companies of the PDL Consortium (Broadcom,
%Citadel, EMC, Facebook, Google, Hewlett-Packard Labs, Hitachi, Intel,
%Microsoft Research, MongoDB, NetApp, Oracle, Samsung America, Seagate,
%Tintri, Two Sigma, Uber, Veritas and Western Digital).
This work was supported in part by the U.S. Army Research Office under grant number W911NF0910273.

>>>>

			\hline
			{\tt map\_basic}
			& 2,1 & \ETAdag{10d 17h} & \ETAdag{16388977} \\
			%[JOB 15] progress: 70330/16388977 brs (0.429130%), ETA 10d 16h 34m 15s (elapsed 43m 21s)
			%[JOB 15] TIMED OUT (0.429130%; ETA 10d 16h 34m 15s)
			\hline
			{\tt map\_basicer}
			& 2,1 & \cpu{877.44}& \ints{28635}	\\
			& 2,2 & \ETAdag{2d 7h} & \ETAdag{5925634} \\
			& 3,1 & \ETAdag{468d 13h} & \ETAdag{35893653} \\
			% [JOB 5] progress: 100181/35893653 brs (0.279105%), ETA 1y 103d 12h 50m 47s (elapsed 59m 24s)
			%[JOB 5] TIMED OUT (0.279105%; ETA 1y 103d 12h 50m 47s)
			% i don't know what this is but it was hanging out under 3,1 here :<
			%[JOB 6] progress: 35042/5925634 brs (0.591363%), ETA 2d 6h 40m 30s (elapsed 21m 12s)
			%[JOB 6] TIMED OUT (0.591363%; ETA 2d 6h 40m 30s)
		\end{tabular}
	\end{center}
	\caption{Transactional tests verified (or not) by Landslide.}
	\label{tab:verifs}
\end{table}

Several larger tests we were unable to complete before the submission deadline are also shown.
%Their completion times are listed as {\em unk.}
These are indicated with $\dagger$ %\ETAdag{}
and their listed ETAs and state space sizes represent Landslide's estimate %at the moment of timeout.
after a timeout of 1 hour
(hopefully enough for data-race PPs to saturate and estimates to stablize,
although note the two estimate types use different algorithms so may disagree significantly).
%For these, Table~\ref{tab:timeouts} presents their size estimates at the moment of timeout.
Though these tests proved beyond our reach,
in contrast with the instant gratification expected of bug-finding,
%we believe
the verification guarantee may in some cases be worth the estimated time requirement
for widely-used industrial implementations.
Future work may also expand our coverage upon this frontier \cite{landslide-phdthesis}.

@@

We noticed significant slowdowns when varying several aspects of our experimental environment.
For example, multiple Landslides running at once
%on our test machine
slow each other down,
%(including multiple single state spaces in a single test when not using {\tt -M}),
likely arising from kernel resource contention as Landslide uses {\tt fork()} to save simulation state.
Table~\ref{tab:variants} shows the impact of running a single Landslide instance with {\tt -M} on
{\tt counter}(2,2) %(384 distinct interleavings)
with various programs also running, despite never saturating our test system's 8 CPUs.
% for i in `seq 1 10`; do ./landslide -X -v -p counter -M 2>&1 | grep 'total CPU time'; done

Table~\ref{tab:variants2} shows
the impact of Landslide needing to automatically annotate {\tt counter}(2,2) being run for the first time,
rather than reusing existing instrumentation,
as well as the variance of Quicksand mode. %(i.e., without {\tt -M}).
Because Quicksand repeats more work across jobs,
rather than comparing it to a baseline,
we show in the right column how many total interleavings each approach executed.
(The maximal state space alone comprises 384 in all cases.)
%(2128 interleavings total).
The {\sf quicksand} variance was surprisingly bimodal,
with 6 samples in a distribution of 106.14$\pm$1.22 and 4 in 142.11$\pm$1.67,
suggesting two distinct scheduling patterns for its workqueue threads.
Future work should figure out why.

@@

Many such branches may be equivalent;
for example,
explicit and conflict aborts need not be tested separately in transactions
whose failure paths do not distinguish the cause,
and the perennial {\tt \_XABORT\_RETRY} abort can be skipped entirely
if the client abstracts it away behind a retry loop.
In fact, applying both reductions simultaneously
amounts to the STM concurrency model \cite{stm-pldi06}%
\footnote{Proof left to future work. We've proved enough here already.}
and may reduce the state space even smaller than the original.
Testing them by hand
\footnote{Using visual inspection of the test to trust the reductions' soundness.}
reduced {\tt map\_basicer}(2,1)'s 28635 interleavings to 11577,
and produced the same 384-interleaving state space on {\tt counter}(2,2)
(in which DPOR triggers conflict aborts on every transaction anyway).
%and {\tt map\_basic}(2,1)'s 1-hour size estimate from 16388977 to 16126960. % about the same after all...
%(although {\tt map\_basic}(2,1) lies still beyond reach).
%We tested both reductions by hand, trusting their soundness by visual inspection,
%on {\tt map\_basic}(2,1),
%and preview the resulting state space sizes in Table~\ref{tab:abort-codes}.
%
%\begin{table}[h]
%	\caption{...}
%	\label{tab:abort-codes}
%\end{table}
%
Future work could use static or dynamic flow analysis to
identify such reduction opportunities automatically;
for now, this feature is disabled by default but accessible via the {\tt -A} command-line option
%to Landslide
(in addition to {\tt -X}).

{\bf Relaxed memory orderings.}
Section~\ref{sec:design}'s formalization of thread interleavings does not account for read/write reorderings
possible on relaxed consistency architectures \cite{memory-consistency-models}.
In fact,
even after \cite{htm-mario}'s proposed fix,
our running example program is still incorrect on Total Store Order (TSO) architectures such as x86.
Despite stores being totally-ordered, x86 may still reorder stores after subsequent loads.
Accordingly, an execution of $8,9a,9b,9c$ in Figure~\ref{fig:htm-fixed}
may be locally visible to another thread as $9a,8,9b,9c$,
and hence an apparent interleaving of
\[
	\tiat{1},\tjat{1-5},\tiat{7},\underline{\tiat{9a}},\tkat{1-5},\underline{\tiat{8}},\tiat{9b-B}
\]
is possible
(reordered accesses underlined for emphasis).
An {\tt mfence} barrier is needed between lines 8 and 9 to solve this problem on TSO \cite{tsx-need-barrier}.
On Partial Store Order (PSO) architectures, even more barriers may be necessary.

@@

We thank Mario Dehesa-Azuara for generously providing the HTM data structures and benchmarks used in our evaluation.
%We also thank the member companies of the PDL Consortium (Broadcom,
%Citadel, EMC, Facebook, Google, Hewlett-Packard Labs, Hitachi, Intel,
%Microsoft Research, MongoDB, NetApp, Oracle, Samsung America, Seagate,
%Tintri, Two Sigma, Uber, Veritas and Western Digital).
This work was supported in part by the U.S. Army Research Office under grant number W911NF0910273.
\vspace{1em}
